---
title: Analyzing and Predicting Bombus Latreille’s Distribution Across North America
  Using Hierarchical Bayesian Models
author: "Paulina Brown"
date: "2024-09-20"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Load packages
library(bayesrules)
library(tidyverse)
library(rstanarm)
library(bayesplot)
library(tidybayes)
library(broom.mixed)
library(lubridate)
library(ggplot2)
library(dplyr)

# Load data
species_data <- read.csv("/Users/paulinabrownp/Desktop/Research Project/BombusLatreille.csv")
temperature_data <- read.csv("/Users/paulinabrownp/Desktop/Research Project/averagetemperature.csv")
plant_data <- read.csv("/Users/paulinabrownp/Desktop/Research Project/Monarda.csv")
```

# Data Preprocessing
## Preprocess Temperature Data
```{r}
temperature_data <- rename(temperature_data, Month = Year)
colnames(temperature_data) <- as.character(colnames(temperature_data))

temperature_long <- pivot_longer(temperature_data, 
                                 cols = matches("^X\\d{4}$"), 
                                 names_to = "Year", 
                                 values_to = "Temperature")

# Convert 'Year' to numeric after stripping prefixes
temperature_long$Year <- as.numeric(sub("^X", "", temperature_long$Year))

# Filter for years 2013 to 2023
temperature_filtered <- filter(temperature_long, Year >= 2013 & Year <= 2023)

temperature_filtered$Year <- as.numeric(as.character(temperature_filtered$Year))

# Remove rows with NAs in the Temperature column from temperature_filtered
temperature_filtered <- temperature_filtered %>% filter(!is.na(Temperature))

# Map 'Code' from CAN, MEX, USA to CA, MX, US
country_code_map <- c(CAN = "CA", MEX = "MX", USA = "US")
temperature_filtered$countryCode <- country_code_map[temperature_filtered$Code]

# Filter temperature data for North American countries
north_american_countries <- c("US", "CA", "MX")
temperature_filtered <- temperature_filtered %>%
  filter(countryCode %in% north_american_countries)
```

## Preprocess Species Data
```{r}
species_data <- rename(species_data, Year = year)
colnames(species_data) <- as.character(colnames(species_data))

species_filtered <- species_data %>%
  mutate(Year = as.numeric(year(ymd_hms(dateIdentified)))) %>%
  filter(Year >= 2013 & Year <= 2023) %>%
  filter(!is.na(stateProvince) & stateProvince != "") %>%
  mutate(Presence = ifelse(occurrenceStatus == "PRESENT", 1, 0))
```

## Preprocess Plant Data
```{r}
plant_data <- rename(plant_data, Year = year)
colnames(plant_data) <- as.character(colnames(plant_data))

plant_filtered <- plant_data %>%
  mutate(Year = as.numeric(year(ymd_hms(dateIdentified)))) %>%
  filter(Year >= 2013 & Year <= 2023) %>%
  filter(!is.na(stateProvince) & stateProvince != "") %>%
  mutate(Presence = ifelse(occurrenceStatus == "PRESENT", 1, 0))
```

# Data Agreggation and Merging
```{r}
# Aggregate temperature data by year
temperature_aggregated <- temperature_filtered %>%
  group_by(Year, countryCode) %>%
  summarise(Temperature = mean(Temperature, na.rm = TRUE))

# Aggregate plant data by Year and stateProvince
plant_aggregated <- plant_filtered %>%
  group_by(Year, stateProvince) %>%
  summarise(PlantCount = n(), .groups = 'drop')

# Merge temperature and species data
species_temperature <- left_join(species_filtered, temperature_aggregated, by = c("Year", "countryCode"))

# Merge temperature and plant data
plant_temperature <- left_join(plant_filtered, temperature_aggregated, by = c("Year", "countryCode"))

species_count_data <- species_temperature %>%
  group_by(stateProvince, Year) %>%
  summarise(
    SpeciesCount = sum(Presence, na.rm = TRUE),
    Temperature = mean(Temperature, na.rm = TRUE),
    .groups = "drop"
  )

plant_count_data <- plant_temperature %>%
  group_by(stateProvince, Year) %>%
  summarise(
    PlantCount = sum(Presence, na.rm = TRUE),
    Temperature = mean(Temperature, na.rm = TRUE),
    .groups = "drop"
  )

# Convert stateProvince to UTF-8 encoding. 
plant_count_data$stateProvince <- iconv(plant_count_data$stateProvince, from = "Windows-1252", to = "UTF-8")
# Manually correct the remaining misencoded names
plant_count_data$stateProvince <- gsub("Michoac‡n", "Michoacán", plant_count_data$stateProvince)
plant_count_data$stateProvince <- gsub("Nuevo Le—n", "Nuevo León", plant_count_data$stateProvince)
plant_count_data$stateProvince <- gsub("QuŽbec", "Québec", plant_count_data$stateProvince)
```
# Exploratory Data Analysis
```{r}
# Exploring Temperature distribution across years
ggplot(temperature_aggregated, aes(x = Year, y = Temperature)) +
  geom_line(group=1, color="blue") +
  geom_point(color="red") +
  labs(title = "Average Temperature Over Time", x = "Year", y = "Temperature")

# Species count over time
ggplot(species_count_data, aes(x = Year, y = SpeciesCount)) +
  geom_line(group=1, color="green") +
  geom_point(color="darkgreen") +
  labs(title = "Species Count Over Time", x = "Year", y = "Species Count")

# Plant count over time
ggplot(plant_count_data, aes(x = Year, y = PlantCount)) +
  geom_line(group=1, color="mediumorchid") +
  geom_point(color= "darkmagenta") +
  labs(title = "Plant Count Over Time", x = "Year", y = "Plant Count")

# Relationship between Temperature and Species Count
ggplot(species_count_data, aes(x = Temperature, y = SpeciesCount)) +
  geom_point() +
  geom_smooth(method = "loess", color = "red") +
  labs(title = "Temperature vs Species Count", x = "Temperature", y = "Species Count")

# Relationship between Temperature and Plant Count
ggplot(plant_count_data, aes(x = Temperature, y = PlantCount)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +
  labs(title = "Temperature vs Plant Count", x = "Temperature", y = "Plant Count")
```
# Model Development: Hierarchical Bayesian Model
## Poisson Models
```{r}
# Hierarchical Model for Species Count
species_model <- stan_glmer(
  SpeciesCount ~ Temperature + (1 | stateProvince/Year),
  data = species_count_data,
  family = poisson(link = "log"),  # Assuming count data follows a Poisson distribution
  prior = normal(0, 1, autoscale = TRUE),  # Normal priors for the coefficients
  chains = 4,
  iter = 2000,
  seed = 123
)

# Hierarchical Model for Plant Count
plant_model <- stan_glmer(
  PlantCount ~ Temperature + (1 | stateProvince/Year),
  data = plant_count_data,
  family = poisson(link = "log"),  # Assuming count data follows a Poisson distribution
  prior = normal(0, 1, autoscale = TRUE),  # Normal priors for the coefficients
  chains = 4,
  iter = 2000,
  seed = 123
)
```

## Negative Binomial Models
```{r}
# Adjusting Species Count Model to Negative Binomial
species_model_nb <- stan_glmer(
  SpeciesCount ~ Temperature + (1 | stateProvince/Year),
  data = species_count_data,
  family = neg_binomial_2(link = "log"),
  prior = normal(0, 1, autoscale = TRUE),
  chains = 4,
  iter = 2000,
  seed = 123
)

# Adjusting Plant Count Model to Negative Binomial
plant_model_nb <- stan_glmer(
  PlantCount ~ Temperature + (1 | stateProvince/Year),
  data = plant_count_data,
  family = neg_binomial_2(link = "log"),
  prior = normal(0, 1, autoscale = TRUE),
  chains = 4,
  iter = 2000,
  seed = 123
)
```

# Posterior Predictive Checks
```{r}
# Posterior predictive check for the Species Count Model
ppc_species <- pp_check(species_model_nb, nsamples = 100, type = "hist") + 
  ggtitle("Posterior Predictive Check for Species Count Model")

# Posterior predictive check for the Plant Count Model
ppc_plant <- pp_check(plant_model_nb, nsamples = 100, type = "hist") + 
  ggtitle("Posterior Predictive Check for Plant Count Model")

# Print the plots to visualize them
print(ppc_species)
print(ppc_plant)

# Posterior predictive check for the adjusted Species Count Model
ppc_species_nb <- pp_check(species_model_nb, nsamples = 100, type = "hist") + 
  ggtitle("Posterior Predictive Check for Adjusted Species Count Model")

# Posterior predictive check for the adjusted Plant Count Model
ppc_plant_nb <- pp_check(plant_model_nb, nsamples = 100, type = "hist") + 
  ggtitle("Posterior Predictive Check for Adjusted Plant Count Model")

# Print the plots
print(ppc_species_nb)
print(ppc_plant_nb)
```

# Implementing Log Transformation to stabilize and normalize distribution
```{r}
# Adding a small constant and applying log transformation
species_count_data$LogSpeciesCount <- log(species_count_data$SpeciesCount + 1)
plant_count_data$LogPlantCount <- log(plant_count_data$PlantCount + 1)


# Hierarchical Model for Log-Transformed Species Count
species_model_log <- stan_glmer(
  LogSpeciesCount ~ Temperature + (1 | stateProvince/Year),
  data = species_count_data,
  family = gaussian(),  # Using Gaussian family for the transformed counts
  prior = normal(0, 1, autoscale = TRUE),
  chains = 4,
  iter = 2000,
  seed = 123
)

# Hierarchical Model for Log-Transformed Plant Count
plant_model_log <- stan_glmer(
  LogPlantCount ~ Temperature + (1 | stateProvince/Year),
  data = plant_count_data,
  family = gaussian(),
  prior = normal(0, 1, autoscale = TRUE),
  chains = 4,
  iter = 2000,
  seed = 123
)

# Posterior predictive check for the log-transformed Species Count Model
ppc_species_log <- pp_check(species_model_log, nsamples = 100, type = "hist") + 
  ggtitle("Posterior Predictive Check for Log-Transformed Species Count Model")

# Posterior predictive check for the log-transformed Plant Count Model
ppc_plant_log <- pp_check(plant_model_log, nsamples = 100, type = "hist") + 
  ggtitle("Posterior Predictive Check for Log-Transformed Plant Count Model")

# Print the plots
print(ppc_species_log)
print(ppc_plant_log)
```

# Predictions and Visualizations
```{r}
# Split data into training and validation sets (70% train, 30% test)
set.seed(123)  # for reproducibility
training_indices <- sample(1:nrow(species_data), 0.7 * nrow(species_data))
species_train <- species_count_data[training_indices, ]
species_test <- species_count_data[-training_indices, ]

# Fit hierarchical model on training data
species_model <- stan_glmer(
  SpeciesCount ~ Temperature + (1 | stateProvince/Year),
  data = species_train,
  family = poisson(link = "log"),
  prior = normal(0, 1, autoscale = TRUE),
  chains = 4,
  iter = 2000,
  seed = 123
)
```

# Prediction and Performance Analysis
```{r}
# Residual Analysis and Diagnostics
# Calculate residuals manually
species_residuals <- residuals(species_model, type = "response")  # Response residuals
species_fitted <- fitted(species_model)  # Fitted values

# Create a residuals vs. fitted values plot
ggplot(data = NULL, aes(x = species_fitted, y = species_residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

# Check for normality of residuals using a histogram
ggplot(data = NULL, aes(x = species_residuals)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  geom_density(color = "red", size = 1) +
  labs(title = "Histogram of Residuals",
       x = "Residuals",
       y = "Density") +
  theme_minimal()

# Optional: Perform a Q-Q plot for residuals
qqnorm(species_residuals, main = "Q-Q Plot of Residuals")
qqline(species_residuals, col = "red")

# Convergence Diagnostics
# Trace plots for parameters
posterior <- as.array(species_model)
mcmc_trace(posterior, pars = c("(Intercept)", "Temperature"), 
           facet_args = list(labeller = label_parsed)) +
  labs(title = "Trace Plots for Species Model Parameters")

# Calculate R-hat values
# Access R-hat values directly using rhat() function
species_rhat <- rhat(species_model)  # Extract R-hat values for all parameters
print(species_rhat)

# Posterior Predictive Checks
pp_check(species_model, type = "dens_overlay") +
  ggtitle("Posterior Predictive Check for Species Count Model")

# Predictive Performance Evaluation
# Predict on the test dataset (if available)
species_test <- species_count_data  # Replace with your actual test dataset
predictions <- posterior_predict(species_model, newdata = species_test)

# Summarize predictions (e.g., take the mean of posterior draws for each observation)
species_test$predicted_counts <- apply(predictions, 2, mean)  # Column-wise mean

# Compare predictions to actual values
ggplot(species_test, aes(x = predicted_counts, y = SpeciesCount)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, col = "red") +  # Perfect fit line
  labs(title = "Predicted vs Actual Species Count",
       x = "Predicted Species Count",
       y = "Actual Species Count") +
  theme_minimal()

# Simple example of 5-fold cross-validation
cv_errors <- numeric(5)
folds <- cut(seq(1, nrow(species_count_data)), breaks = 5, labels = FALSE)

for(i in 1:5) {
  test_indices <- which(folds == i)
  train_indices <- setdiff(seq(1, nrow(species_count_data)), test_indices)
  model <- stan_glmer(
    SpeciesCount ~ Temperature + (1 | stateProvince/Year),
    data = species_count_data[train_indices, ],
    family = poisson(link = "log"),
    prior = normal(0, 1, autoscale = TRUE),
    chains = 4,
    iter = 2000,
    seed = 123
  )
  predictions <- posterior_predict(model, newdata = species_count_data[test_indices, ])
  cv_errors[i] <- sqrt(mean((rowMeans(predictions) - species_count_data$SpeciesCount[test_indices])^2))  # RMSE
}

mean(cv_errors)  # Average RMSE over all folds

# Generate a final visualization
ggplot(species_count_data, aes(x = Year, y = SpeciesCount)) +
  geom_line(aes(group = stateProvince), color = "blue") +
  geom_point(aes(color = stateProvince)) +
  labs(title = "Species Count Over Time by Province",
       x = "Year",
       y = "Species Count") +
  theme_minimal()
```
